Parameter Recovery Validation

### 2025-05-30 nbs/concepts/parameter-recovery-validation/index.qmd
docs: generate sample data for parameter recovery validation
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

We generate synthetic datasets representing each of the four gene expression patterns using the generic predictive sampling interface:

```python
# Generate validation datasets for all patterns using generic API
validation_datasets = {}
validation_parameters = {}

# Define pattern configurations (2 parameter sets per pattern = 8 total)
pattern_configs = [
    # Activation patterns
    {"pattern": "activation", "set_id": 1, "seed": 42},
    {"pattern": "activation", "set_id": 2, "seed": 43},
    # Decay patterns
    {"pattern": "decay", "set_id": 1, "seed": 44},
    {"pattern": "decay", "set_id": 2, "seed": 45},
    # Transient patterns
    {"pattern": "transient", "set_id": 1, "seed": 46},
    {"pattern": "transient", "set_id": 2, "seed": 47},
    # Sustained patterns
    {"pattern": "sustained", "set_id": 1, "seed": 48},
    {"pattern": "sustained", "set_id": 2, "seed": 49},
]

for config in pattern_configs:
    pattern = config["pattern"]
    set_id = config["set_id"]
    seed = config["seed"]
    dataset_key = f"{pattern}_{set_id}"

    print(f"Generating synthetic data for {dataset_key}...")

    # Set PyTorch random seed for reproducibility
    torch.manual_seed(seed)
    pyro.set_rng_seed(seed)

    # Sample system parameters constrained to this pattern
    true_system_params = model.sample_system_parameters(
        pattern=pattern,
        set_id=set_id,
        num_samples=1,
        constrain_to_pattern=True,
        n_genes=5,
        n_cells=200
    )

    # Generate synthetic data using the generic predictive sampling interface
    torch.manual_seed(seed + 100)
    pyro.set_rng_seed(seed + 100)
    adata = model.generate_predictive_samples(
        num_cells=200,
        num_genes=5,
        samples=true_system_params,  # Use constrained parameters as "posterior"
        return_format="anndata"
    )

    # Store dataset and true parameters
    validation_datasets[dataset_key] = adata
    validation_parameters[dataset_key] = true_system_params

    print(f"  Generated {adata.n_obs} cells × {adata.n_vars} genes")
    print(f"  Pattern: {adata.uns.get('pattern', 'unknown')}")
    print(f"  True parameters: {list(true_system_params.keys())}")

print(f"\nGenerated {len(validation_datasets)} synthetic datasets")
```

>>>>

We generate synthetic datasets representing each of the four gene expression patterns using the new streamlined validation API:

```python
# Generate validation datasets for all patterns using the new API
print("📊 Generating validation datasets...")

validation_datasets = model.generate_validation_datasets(
    patterns=['activation', 'decay', 'transient', 'sustained'],
    n_genes=5,
    n_cells=200,
    n_parameter_sets=2,  # 2 parameter sets per pattern = 8 total datasets
    seed=42
)

print(f"✅ Generated {len(validation_datasets)} validation datasets")

# Display dataset information
for dataset_key, dataset_info in validation_datasets.items():
    adata = dataset_info['adata']
    pattern = dataset_info['pattern']
    true_params = dataset_info['true_parameters']

    print(f"\n{dataset_key}:")
    print(f"  Data shape: {adata.shape}")
    print(f"  Pattern: {pattern}")
    print(f"  True parameters: {list(true_params.keys())}")
    print(f"  Metadata: {dataset_info.get('metadata', {})}")

print(f"\n🎯 Ready for parameter recovery validation!")
```
